An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-04-19 conclusion.tex
Updates from Overleaf
introduced logic macros

Propositional linear temporal logic (LTL) and automata are two common specification formalisms for software and hardware systems. While temporal logic has a more declarative
flavor, automata are more operational, describing how the specified system
progresses. There has been several proposed extensions to LTL that extend
its expressive power to that of related automata formalisms. We proposed here
a simple extension for propositional LTL that is based on adding propositions that summarize the prefix of an execution. Conceptually, this extension puts the specification in between propositional LTL and automata, as the additional
variables can be seen as representing the state of an automaton
that is synchronized with the temporal property. It is shown to have
the same expressive power as automata models, and is in particular
appealing for runtime verification of past (i.e., safety) temporal properties,
which already are based on summarizing the value of subformulas over
observed prefixes.

We demonstrated that first-order linear temporal logic (FLTL), which
can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.
We showed that for the first-order case, unlike the propositional case,
this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.

We presented a monitoring
algorithm for propositional past time temporal logic, extending a classical algorithm, and similarly presented an algorithm for 
first-order past temporal logic.
Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results.

>>>>

Propositional linear temporal logic (LTL) and automata are two common specification formalisms for software and hardware systems. While temporal logic has a more declarative
flavor, automata are more operational, describing how the specified system
progresses. There has been several proposed extensions to LTL that extend
its expressive power to that of related automata formalisms. We proposed here
a simple extension for propositional LTL that is based on adding propositions that summarize the prefix of an execution. Conceptually, this extension puts the specification in between propositional LTL and automata, as the additional
variables can be seen as representing the state of an automaton
that is synchronized with the temporal property. It is shown to have
the same expressive power as automata models, and is in particular
appealing for runtime verification of past (i.e., safety) temporal properties,
which already are based on summarizing the value of subformulas over
observed prefixes.
%
We demonstrated that first-order linear temporal logic (\FLTL{}), which
can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.
We showed that for the first-order case, unlike the propositional case,
this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.
%
We presented a monitoring
algorithm for propositional past time temporal logic, extending a classical algorithm, and similarly presented an algorithm for 
first-order past temporal logic.
Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results.
%
Future work includes performing additional experiments, and making further comparisons to other formalisms. Finally we would like to study further extensions, exploring the space between logic and programming.

### 2019-04-19 implementation.tex
introduced logic macros

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
%
The ASCII version of the logic uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

>>>>

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)\FLTL{}.
These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
%
The ASCII version of the logic uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

### 2019-04-19 main.tex
introduced logic macros

Hence, the sequence of states 
\[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\]
is a model of this formula, while 
\[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \]
is not.
The QFLTL formula is a bit involved, hence given in several parts (macros), where the main formula appears last.

@@

\iffalse
We prove that the above property cannot be expressed in EPFLTL.
We use a combinatorial argument, showing that
the information that is available in any EPFLTL
property
after a prefix is not sufficient to distinguish between two models (executions), one satisfying and one not satisfying it.

In EPFLTL, after every prefix of the input, it is
sufficient to keep a finite set of (non temporal) relations over the
data elements seen in the input, which
summarizes the prefix. This follows from the set
semantics, which appears in Section~\ref{EPLTLRV}, 
where the interpretation of
each subformula, and the auxiliary relations,
at a particular state is a relation, and the interpretation 
of a property a prefix of length $i$ only depends on
these relations. Now, the amount of information
that summarizes an observed prefix for
EPFLTL is is polynomial
in $N$, consisting of a finite number
of relations over
$N$ elements. The number of different tuples
for such relations can be shown, inductively, to
be polynomial in $N$ (beach parameter
of each relation can be one out of the $N$ possibilities, or a representation the values not seen yet).
However, the above property depends on whether
a set of values that appear as parameters to
the relation $v$, between two consecutive appearances
of $r$'s, are the same. For that, one needs to keep some representation of all the
subsets of the elements that appeared in the input so far; by contradiction, any summary representation of the observed prefix that
would identify two prefixes with some different sets, would not be able to distinguish between
sequences that extend this prefix with the set
of elements that appears in one prefix but not the other.
But the number of such subsets is exponential in $N$. \qed
\fi

@@

\iffalse
{\bf Comments.} First, note that the reason that QFLTL {\em can} express the property in Theorem~\ref{theo2}, is that it allows to augment the input relations with auxiliary relations that are not fixed. That is, for an auxiliary relation $R$,
$R [ i ]$ is not fixed, and can change per different
prefixes (of length at least $i$), or even in different
subformulas. Second, given a simple extension to EFLTL, where 
one can use a variable that represents an updatable
counter (or a generator of distinct values), one can express the property used in~\ref{theo2}, by
keeping a relation where values between different occurrences of the event $r$ are paired with distinct values, hence, representing an unbounded set of sets.
\fi

%%%M
\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
One can translate any QLTL property into a second-order monadic logic in a standard
way~\cite{Thomas}, and then from
second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set. A muller automaton 
accepts a run if it 
passes infinitely often in exactly
one of the specified accepting sets. To formulate an equivalent ELTL formula,
the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
Then we translate $M$ into an ELTL formula
of the form:

>>>>

Hence, the sequence of states 
\[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\]
is a model of this formula, while 
\[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \]
is not.
The \QFLTL{} formula is a bit involved, hence given in several parts (macros), where the main formula appears last.

@@

\iffalse
We prove that the above property cannot be expressed in \EPFLTL{}.
We use a combinatorial argument, showing that
the information that is available in any \EPFLTL{}
property
after a prefix is not sufficient to distinguish between two models (executions), one satisfying and one not satisfying it.

In \EPFLTL{}, after every prefix of the input, it is
sufficient to keep a finite set of (non temporal) relations over the
data elements seen in the input, which
summarizes the prefix. This follows from the set
semantics, which appears in Section~\ref{EPLTLRV}, 
where the interpretation of
each subformula, and the auxiliary relations,
at a particular state is a relation, and the interpretation 
of a property a prefix of length $i$ only depends on
these relations. Now, the amount of information
that summarizes an observed prefix for
\EPFLTL{} is is polynomial
in $N$, consisting of a finite number
of relations over
$N$ elements. The number of different tuples
for such relations can be shown, inductively, to
be polynomial in $N$ (beach parameter
of each relation can be one out of the $N$ possibilities, or a representation the values not seen yet).
However, the above property depends on whether
a set of values that appear as parameters to
the relation $v$, between two consecutive appearances
of $r$'s, are the same. For that, one needs to keep some representation of all the
subsets of the elements that appeared in the input so far; by contradiction, any summary representation of the observed prefix that
would identify two prefixes with some different sets, would not be able to distinguish between
sequences that extend this prefix with the set
of elements that appears in one prefix but not the other.
But the number of such subsets is exponential in $N$. \qed
\fi

@@

\iffalse
{\bf Comments.} First, note that the reason that \QFLTL{} {\em can} express the property in Theorem~\ref{theo2}, is that it allows to augment the input relations with auxiliary relations that are not fixed. That is, for an auxiliary relation $R$,
$R [ i ]$ is not fixed, and can change per different
prefixes (of length at least $i$), or even in different
subformulas. Second, given a simple extension to \EFLTL{}, where 
one can use a variable that represents an updatable
counter (or a generator of distinct values), one can express the property used in~\ref{theo2}, by
keeping a relation where values between different occurrences of the event $r$ are paired with distinct values, hence, representing an unbounded set of sets.
\fi

%%%M
\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
One can translate any \QLTL{} property into a second-order monadic logic in a standard
way~\cite{Thomas}, and then from
second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set. A muller automaton 
accepts a run if it 
passes infinitely often in exactly
one of the specified accepting sets. To formulate an equivalent \ELTL{} formula,
the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
Then we translate $M$ into an \ELTL{} formula
of the form:
